\chapter{Distribution checklist} 

\section{Building Phycas on Ubuntu 10.04}

\begin{enumerate}
\item Prepare operating system for building phycas
\begin{verbatim}
sudo apt-get install g++
sudo apt-get install python-dev
sudo apt-get install autoconf
sudo apt-get install subversion
sudo apt-get install curl
sudo apt-get install git-core
sudo apt-get install xsel
\end{verbatim}
%
\item Download \pathname{ncl-2.1.13.tar.gz} from \url{http://sourceforge.net/projects/ncl/}
\begin{verbatim}
cd $HOME
tar zxvf ncl-2.1.13.tar.gz
cd ncl-2.1.13
./configure
make
sudo make install
\end{verbatim}
Alternatively, download and build NCL from SourceForge Subversion checkout
\begin{verbatim}
cd $HOME
mkdir ndev
cd ndev
mkdir v2.1
svn co https://ncl.svn.sourceforge.net/svnroot/ncl/branches/v2.1 v2.1
cd v2.1
sh ./bootstrap.sh
./configure
make
sudo make install
\end{verbatim}
%
\item Download \pathname{boost\_1\_42\_0.tar.gz} from \url{http://www.boost.org/}
\begin{verbatim}
cd $HOME
tar zxvf boost_1_42_0.tar.gz
\end{verbatim}
%
\item Build bjam executable.
The following should create a new directory 
containing the bjam executable. 
The name of the directory will depend on whether your version of g++ is configured
to build for 32 or 64 bits. 
The directory created should be:\\
\pathname{\$HOME/boost\_1\_42\_0/tools/jam/src/bin.linuxx86}\\
or \\
\pathname{\$HOME/boost\_1\_42\_0/tools/jam/src/bin.linuxx86\_64} \\
con):
\begin{verbatim}
cd $HOME/boost_1_42_0/tools/jam/src
./build.sh
\end{verbatim}
%
\item Download and build phycas
\begin{verbatim}
cd $HOME
mkdir pdev
cd pdev
git clone git@github.com:mtholder/Phycas.git
cd Phycas
\end{verbatim}
%
\item Create file \pathname{\$HOME/pdev/Phycas/go.sh} to contain the following text:
\begin{verbatim}
#!/bin/bash

export OSTYPE="linux"
export BOOST_ROOT="$HOME/boost_1_42_0"
export PATH="${PATH}:$BOOST_ROOT/tools/jam/src/bin.linuxx86"
export BOOST_BUILD_PATH="${BOOST_ROOT}/tools/build/v2"
export PHYCAS_ROOT="$HOME/pdev/trunk"
export NCL_INSTALL_DIR="/usr/local"
export NCL_ALREADY_INSTALLED=1
bjam release
\end{verbatim}

Note that if building bjam created a 64 bit version then you will have to use
\begin{verbatim}
export PATH="${PATH}:$BOOST_ROOT/tools/jam/src/bin.linuxx86_64"
\end{verbatim}
to setup the correct path

%
\item Make \pathname{\$HOME/pdev/Phycas/go.sh} executable and build phycas
\begin{verbatim}
cd $HOME/pdev/trunk
chmod +x go.sh
./go.sh
\end{verbatim}
%
\item Prepare libraries for dynamic loading
\begin{verbatim}
/sbin/ldconfig $HOME/pdev/Phycas/phycas/Conversions /usr/local/lib/ncl
\end{verbatim}
(you may have to run this via sudo).
%
\item Create tar file of phycas distribution:
\begin{verbatim}
cd $HOME/pdev/trunk
python build_phycas_tarball.py
\end{verbatim}
%
\item Locate the python dist-packages directory:
\begin{verbatim}
python -c "from distutils.sysconfig import get_python_lib; print get_python_lib()"
\end{verbatim}
%
\item Assuming previous step revealed that the python dist-packages directory is \pathname{/usr/lib/python2.6/dist-packages}, move the tar file there and unpack:
\begin{verbatim}
cd /usr/lib/python2.6/dist-packages
sudo tar xvf "$HOME/pdev/trunk/phycas_distr.tar"
\end{verbatim}
%
\item See if phycas can be loaded into python:
\begin{verbatim}
python
>>> from phycas import *
\end{verbatim}

\subsection*{Preparing source distribution for the \url{phycas.org} web site}
The following command will create a \pathname{dist} directory inside of which is a \pathname{Phycas-1.2.0.tar.gz} file. The ``1.2.0'' part is obtained from \pathname{phycasver.txt}, so make sure that is correct first. The shell script first creates MANIFEST.in by concatenating MANIFEST.common and MANIFEST.linux. Subsequently, the distutils sdist command is used to create the distribution including all files listed in MANIFEST.in.
\begin{verbatim}
cd $HOME/pdev/Phycas
./build_linux_source_distr.sh
\end{verbatim}

\subsection*{Making python less painful}
bpython and ipython allow for tab completion and other have other helpful features that make phycas less painful to work with.

\subsubsection*{setuptools}
To install them, start by installing setuptools.  See \url{http://pypi.python.org/pypi/setuptools#cygwin-mac-os-x-linux-other}
This installation  will entail  grabbing the egg from \url{http://pypi.python.org/packages/2.6/s/setuptools/setuptools-0.6c11-py2.6.egg#md5=bfa92100bd772d5a213eedd356d64086}
and then installing it 
\begin{verbatim}
sh setuptools-0.6c11-py2.6.egg
\end{verbatim}
That may require sudo.

Setuptools should give you ``easy\_install'' on your path.
After that you should be able to run
\begin{verbatim}
easy_install bpython
easy_install ipython
\end{verbatim}
to install bpython and iPython.

\section{Building Phycas on Windows 7}

\subsection*{Preparing binary distribution for the \url{phycas.org} web site}
The following command will create a \pathname{dist} directory inside of which is a \pathname{Phycas-1.2.0.zip} file. The ``1.2.0'' part is obtained from \pathname{phycasver.txt}, so make sure that is correct first. The shell script first creates MANIFEST.in by concatenating MANIFEST.common and MANIFEST.windows. Subsequently, the distutils sdist command is used to create the distribution including all files listed in MANIFEST.in.
\begin{verbatim}
cd $HOME/pdev/Phycas
./build_windows_source_distr.sh
\end{verbatim}

\section{Building Phycas on Mac OS 10.6.4 (Snow Leopard)}

\end{enumerate}
